---
title: "Introduction to R: Week 1 Solutions"
author: "Maria Doyle"
date: "`r format(Sys.time(), '%d %B %Y')`"
output: 
  html_notebook:
    toc: yes
    toc_float: yes
    toc_depth: 2
subtitle: visualising RNA-seq data with ggplot2
---


#### Exercise
* Read in the normalised counts (from "/training/r-intro-tidyverse/data/normcounts.tsv.gz")
* How many rows and columns are in the file
* In the first row, what is the sample id, the gene symbol and the value for the counts.
* In the last row, what is the sample id, the gene symbol and the value for the counts.
* Make as many plots as you can from bar plots, box plots, violin plots and density plots using the normalised counts (some starter code is below)
* Make separate plots coloured by Sample, CellType, Status, where appropriate.
* Do you notice any differences with the normalised counts versus the unnormalised counts that we used previously (`counts` dataset).

```{r}
norm_counts <- read_tsv("/training/r-intro-tidyverse/data/normcounts.tsv.gz")
```

```{r}
dim(norm_counts)
```

There are 189,648 rows and 5 columns.


```{r}
head(norm_counts)
```
The column names are: Sample, CellType, Status, SYMBOL,  Norm_counts.

In the first row, the sample is DG, the gene is Xkr4 and the counts value is 3.932532.

```{r}
tail(norm_counts)
```

In the last row, the sample is LF, the gene is Erdr1 and the counts value is 3.0911666.

## Bar plots

Bar plot coloured by Sample.

```{r}
ggplot(data=norm_counts, mapping=aes(x=Sample, weight=Norm_counts, fill=Sample)) +
  geom_bar() 
```

Bar plot coloured by CellType.

```{r}
ggplot(data=norm_counts, mapping=aes(x=Sample, weight=Norm_counts, fill=CellType)) +
  geom_bar() 
```

Bar plot coloured by Status.

```{r}
ggplot(data=norm_counts, mapping=aes(x=Sample, weight=Norm_counts, fill=Status)) +
  geom_bar() 
```


## Box plots

Box plots coloured by Sample.

```{r}
ggplot(data=norm_counts, mapping=aes(x=Sample, y=Norm_counts, fill=Sample)) +
  geom_boxplot() 
```

Box plots coloured by CellType.

```{r}
ggplot(data=norm_counts, mapping=aes(x=Sample, y=Norm_counts, fill=CellType)) +
  geom_boxplot() 
```


Box plots coloured by Status.

```{r}
ggplot(data=norm_counts, mapping=aes(x=Sample, y=Norm_counts, fill=Status)) +
  geom_boxplot() 
```


## Violin plots

Violin plots coloured by Sample.


```{r}
ggplot(data=norm_counts, mapping=aes(x=Sample, y=Norm_counts, fill=Sample)) +
  geom_violin()
```

Violin plots coloured by CellType.


```{r}
ggplot(data=norm_counts, mapping=aes(x=Sample, y=Norm_counts, fill=CellType)) +
  geom_violin()
```


Violin plots coloured by Status.


```{r}
ggplot(data=norm_counts, mapping=aes(x=Sample, y=Norm_counts, fill=Status)) +
  geom_violin()
```

## Density Plots

Density plots coloured by Sample.

```{r}
ggplot(data=norm_counts, mapping=aes(x=Norm_counts, colour=Sample)) +
  geom_density() 
```


Density plots coloured by CellType.

```{r}
ggplot(data=norm_counts, mapping=aes(x=Norm_counts, colour=CellType)) +
  geom_density() 
```

Density plots coloured by Status.

```{r}
ggplot(data=norm_counts, mapping=aes(x=Norm_counts, colour=Status)) +
  geom_density() 
```


Differences versus the unnormalised counts
* The normalised counts density plots don't have the large peak on left that's present in the unnormalised counts (the normalised counts have been filtered to remove lowly expressed genes)
* The normalised counts box plots and violin plots, don't give the warning about non-infinite values (the normalised counts have been filtered to remove lowly expressed genes)
* In the box plots of normalised data, the median line is more similar in the samples compared to the unnormalised data.  