---
title: "Introduction to R: Week 4 Practice"
author: "Maria Doyle"
date: "`r format(Sys.time(), '%d %B %Y')`"
output: 
  html_notebook:
    toc: yes
    toc_float: yes
    toc_depth: 2
subtitle: REDCap challenge
---

#### Acknowledgements
Thanks to Dorothea Lesche, Peter Mac postdoc and R course attendee (May 2019) for the idea.

## Challenge

You've collected some information on patients for your research study using [REDCap](https://www.project-redcap.org/) and now you want to do some analysis of the data. You need to separate the patient information using the information collected. You need one file containing the information for all patients who had a single tumour occurance and one file containing the information for all patients who had tumour recurrances. You also need a file containing the information for recurrance patients who had a tumour recur in the same location.

The aim of this challenge is to separate the REDCap patient information into files based on the information collected.

```{r}
library(tidyverse)
patients <- read_csv("/training/r-intro-tidyverse/data/redcap_patients.csv")
```

This code was supplied.

```{r}
# copy DOB and last_coms down by one row
patients <- fill(patients, DOB, last_coms, .direction = c("down"))

# delete rows where the rep_inst column is NA
patients <- drop_na(patients, rep_inst)
```


```{r}
# count how many times each patient appears (no. of occurances)
patients_counts <- count(patients, StudyID)

# identify the patients that appear once
patients_uniq <- filter(patients_counts, n == 1)

# extract the ids for these patients
patients_uniq_ids <- pull(patients_uniq, StudyID)

# make a file of the patients that have only single occurances
patients_single <- filter(patients, StudyID %in% patients_uniq_ids)

# make a file of the patients that have only single occurances
patients_recur <- filter(patients, !(StudyID %in% patients_uniq_ids))

# write out the single occurance patients
write_csv(patients_single, "redcap_single.csv")

# write out the reccurance patients
write_csv(patients_recur, "redcap_recur.csv")
```

Extra task, identify patients with recurrances in the same location.

```{r}
recur_loc_counts <-  group_by(patients_recur, StudyID, breast_side)
recur_loc_same <-  filter(recur_loc_counts, n() > 1)
```

